---
title: "Data Processing Pipeline"
subtitle: "BCI 50-ha Plot — Tree Architecture × Liana Infestation"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    embed-resources: true
execute:
  eval: false       # code shown but not run — paths are placeholders
  warning: false
  message: false
---

## Overview

This document outlines the key data processing steps that produced the
analysis-ready datasets used in `analysis_and_figures.qmd`. It covers:
field data cleaning and infestation-history classification; matching of
TLS-segmented tree point clouds to the ForestGEO stem map; extraction of
crown architecture metrics using ITSMe; and estimation of plant area index
(PAI) via a voxel-based pipeline.

This is intended as a readable record of the processing workflow for anyone
curious about how the data were prepared — it does not constitute a
fully self-contained reproducible pipeline. In particular, the upstream
inputs (subplot-level TLS point clouds, the BCI dendrometer `.rdata`
timeseries, and individual segmented tree files) are large raw data objects
that are not included in this repository. **Raw data may be made available
upon reasonable request.**

Steps not documented here include initial point cloud co-registration
(performed in RiSCAN Pro), individual tree segmentation (performed with
RayCloudTools), leaf-wood separation (performed with a deep-learning
separator), and manual quality checking and cleaning of segmented trees
in CloudCompare.

---

# 1 Field Data Cleaning

## 1.1 Background

Liana infestation data come from annual dendrometer surveys within 100
non-overlapping 40 × 40 m subplots nested in the BCI 50-ha ForestGEO
plot. Surveys began in 2011 (census 8) and record liana presence on a
five-point scale (0 = no lianas, 1–4 = 25% increments of crown covered).
Additional field assessments of liana climbing mode and infestation
pathway were collected in 2025 for six focal species (≥10 individuals
each). This section cleans both datasets and builds the infestation-history
summaries used in the main analysis.

## 1.2 Setup

```{r}
#| label: setup

library(tidyverse)
library(lubridate)
library(janitor)
library(purrr)
library(stringr)

# ── Paths ──────────────────────────────────────────────────────────────────────
# The BCI dendrometer timeseries is distributed as an .rdata file containing
# a data frame 'bciall' with one row per tree × census visit.
rdata_path <- # insert path to bci01to28proc.rdata (dendrometer timeseries)
field_csv  <- # insert path to fieldwork.csv (2025 climbing-mode survey data)
out_dir    <- # insert path to output folder for processed CSVs

dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

## 1.3 Load datasets

```{r}
#| label: load-raw

# Load the BCI dendrometer .rdata into a temporary environment to avoid
# clobbering the workspace. The object inside is named 'bciall'.
temp_env <- new.env()
load(rdata_path, envir = temp_env)
if (!exists("bciall", envir = temp_env)) stop("bciall not found in .rdata.")
bciall <- get("bciall", envir = temp_env)

# Load the 2025 fieldwork survey (climbing modes, infestation pathway)
fieldwork_raw <- read_csv(field_csv, show_col_types = FALSE) |>
  janitor::clean_names()
```

## 1.4 Clean and deduplicate fieldwork

```{r}
#| label: clean-fieldwork

# Helper: normalise any tree tag to a zero-padded 6-digit character string.
# This ensures reliable joining between datasets that store tags as integers,
# raw strings, or 7-digit formats.
tag6_from <- function(x) {
  xnum <- suppressWarnings(as.integer(x))
  ifelse(is.na(xnum), NA_character_, sprintf("%06d", xnum))
}

fieldwork <- fieldwork_raw |>
  mutate(
    priority = if_else(is.na(priority), NA_character_,
                       str_to_lower(str_squish(priority))),
    tag6 = tag6_from(tag)
  ) |>
  # Normalise adhesive-root spelling variants to a single canonical form
  mutate(across(any_of(c("mode_l1", "five_plus")), ~ {
    v  <- as.character(.)
    v2 <- ifelse(is.na(v), NA_character_, tolower(str_squish(v)))
    ifelse(!is.na(v2) & v2 %in% c("adhesive roots","adhesive_roots","adhesive-roots"),
           "adhesive_roots", v)
  }))

# When a tree appears more than once (e.g. two survey visits), keep the
# most informative row: priority = "completed" first, then most non-NA fields.
resolve_duplicates <- function(df, by = "tag6",
                               prefer_col = "priority",
                               prefer_val = "completed") {
  df |>
    mutate(.non_na_count = rowSums(!is.na(select(., -all_of(by))))) |>
    group_by(across(all_of(by))) |>
    arrange(desc((!!sym(prefer_col)) == prefer_val), desc(.non_na_count)) |>
    slice_head(n = 1) |>
    ungroup() |>
    select(-.non_na_count)
}

fieldwork_dedup <- resolve_duplicates(fieldwork, by = "tag6")
message("Fieldwork rows: ", nrow(fieldwork), " → after dedup: ",
        nrow(fieldwork_dedup))

# Build the cleaned fieldwork table with standardised column names.
# lianas_2019 holds the field-measured liana cover score for 2019;
# infestation_type, mode columns, and notes are retained for H2b analyses.
fieldwork_clean_v2 <- fieldwork_dedup |>
  transmute(
    fid, sbpltnm, tag, species, quadrnt, tag6,
    lianas_2019      = coalesce(lianas, NA_real_),
    notes_2019       = nota,
    notes_2025       = notes_lian,
    liana_tag_1 = tag_1, liana_tag_2 = tag_2, liana_tag_3 = tag_3,
    liana_tag_4 = tag_4, liana_tag_5 = tag_5,
    infestation_type,
    mode_l1, mode_l2, mode_l3, mode_l4, mode_l5,
    five_plus, x, y, priority
  ) |>
  mutate(lianas_2019 = as.numeric(lianas_2019))

write_csv(fieldwork_clean_v2, file.path(out_dir, "fieldwork_clean_v2.csv"))
```

## 1.5 Build annual liana timeseries (wide format)

```{r}
#| label: liana-wide

# Standardise the dendrometer tags and find trees present in both datasets.
bciall2 <- bciall |>
  mutate(tag6 = if_else(!is.na(tag),
                        sprintf("%06d", as.integer(tag)), NA_character_))

# Keep only the dendrometer records for trees that also appear in the
# fieldwork survey — this restricts to the study's focal trees.
bciall_overlap <- bciall2 |>
  filter(!is.na(tag6)) |>
  semi_join(fieldwork_clean_v2 |> filter(!is.na(tag6)) |> distinct(tag6),
            by = "tag6")

# Pivot to wide format: one row per tree, columns lianas_2011 … lianas_2024.
# Where multiple records exist for a single tree × year, the most recent
# non-NA liana score is retained.
yrs   <- 2011:2024L
lcols <- paste0("lianas_", yrs)

bciall_wide <- bciall_overlap |>
  mutate(tag6 = as.character(tag6), date = as.Date(date),
         year = year(date), lianas = as.numeric(lianas)) |>
  filter(!is.na(tag6), year %in% yrs) |>
  arrange(tag6, year, desc(date)) |>
  group_by(tag6, species, year) |>
  summarise(
    lianas = { v <- lianas; idx <- which(!is.na(v))
               if (length(idx)) v[idx[1]] else NA_real_ },
    .groups = "drop") |>
  pivot_wider(id_cols = c(tag6, species),
              names_from = year, values_from = lianas,
              names_prefix = "lianas_")

# Ensure all year columns exist even if no data for that year
for (col in lcols)
  if (!col %in% names(bciall_wide)) bciall_wide[[col]] <- NA_real_
bciall_wide <- bciall_wide |> mutate(across(all_of(lcols), as.numeric))

# Where the dendrometer lacks a 2019 score, fill using the 2025 field
# assessment. The field records liana cover as a percentage (0–100); this
# is rescaled to the 0–4 dendrometer scale by dividing by 25 and rounding.
field_2019_pct <- fieldwork_clean_v2 |>
  filter(!is.na(tag6)) |>
  distinct(tag6, .keep_all = TRUE) |>
  transmute(tag6, lianas_2019_field_pct = as.numeric(lianas_2019))

bciall_wide_filled <- bciall_wide |>
  left_join(field_2019_pct, by = "tag6") |>
  rename(lianas_2019_dendro = lianas_2019) |>
  mutate(
    lianas_2019 = case_when(
      !is.na(lianas_2019_dendro) ~ as.numeric(lianas_2019_dendro),
      !is.na(lianas_2019_field_pct) ~
        pmin(4, pmax(0, round(lianas_2019_field_pct / 25))),
      TRUE ~ NA_real_
    )
  ) |>
  select(-lianas_2019_dendro, -lianas_2019_field_pct)
```

## 1.6 Assign infestation-history categories (H1_category)

```{r}
#| label: infestation-histories

# For H1, trees are classified based on their liana record spanning 2011–2019.
# The classification uses two temporal windows:
#   - up19: 2011–2019 (determines current infestation status)
#   - pre:  2011–2018 (determines data sufficiency and "last lost" year)
#
# Categories:
#   Persistently infested — infested in 2019, first infestation ≤ 2015
#   Recently infested     — infested in 2019, first infestation ≥ 2016
#   Lost lianas (recently lost 2016–2018) — not infested in 2019, last infested ≥ 2016
#   Lost lianas (before 2016)             — not infested in 2019, last infested ≤ 2015
#   Never infested        — no infestation record in 2011–2019, ≥1 pre-2019 survey
#   other/uncertain       — insufficient temporal coverage to classify

build_histories <- function(df, yrs = 2011:2024) {
  lcols <- paste0("lianas_", yrs)
  pre   <- 2011:2018
  up19  <- 2011:2019

  df |>
    mutate(.vals = pmap(list(!!!syms(lcols)), ~ c(...))) |>
    transmute(
      tag6, species, across(all_of(lcols)),
      rec_years_vec  = map(.vals, ~ up19[!is.na(.x[yrs %in% up19])]),
      inf_years_vec  = map(.vals, ~ up19[!is.na(.x[yrs %in% up19]) &
                                           .x[yrs %in% up19] != 0]),
      infested_2019  = map_lgl(.vals, ~ any(!is.na(.x) & .x != 0 &
                                              yrs == 2019)),
      n_rec_pre      = map_int(.vals, ~ sum(!is.na(.x[yrs %in% pre]))),
      any_inf_up19   = map_lgl(inf_years_vec, ~ length(.x) > 0),
      first_inf_u19  = map_int(inf_years_vec, ~ if (length(.x)) min(.x)
                                                else NA_integer_),
      last_inf_pre19 = map_int(.vals, ~ {
        y <- pre[!is.na(.x[yrs %in% pre]) & .x[yrs %in% pre] != 0]
        if (length(y)) max(y) else NA_integer_
      }),
      recorded_years = map_chr(rec_years_vec,
                                ~ if (length(.x)) paste(.x, collapse = ";")
                                  else NA_character_),
      infested_years = map_chr(inf_years_vec,
                                ~ if (length(.x)) paste(.x, collapse = ";")
                                  else NA_character_)
    ) |>
    mutate(
      data_poor   = (n_rec_pre < 2) | recorded_years %in% c("2019","2019;2024"),
      H1_category = case_when(
        data_poor ~ "other/uncertain",
        infested_2019 & !is.na(first_inf_u19) & first_inf_u19 >= 2016 ~
          "Recently infested",
        infested_2019 & !is.na(first_inf_u19) & first_inf_u19 <= 2015 ~
          "Persistently infested",
        !infested_2019 & !is.na(last_inf_pre19) & last_inf_pre19 %in% 2016:2018 ~
          "Lost lianas (recently lost 2016\u20132018)",
        !infested_2019 & !is.na(last_inf_pre19) & last_inf_pre19 <= 2015 ~
          "Lost lianas (before 2016)",
        !any_inf_up19 & n_rec_pre > 0 ~ "Never infested",
        TRUE ~ "other/uncertain"
      )
    ) |>
    select(tag6, species, all_of(lcols),
           recorded_years, infested_years,
           first_inf = first_inf_u19,
           last_inf  = last_inf_pre19,
           H1_category)
}

bciall_infestation_histories <- build_histories(bciall_wide_filled, yrs)

message("H1 category distribution:")
print(count(bciall_infestation_histories, H1_category, sort = TRUE))

write_csv(bciall_infestation_histories,
          file.path(out_dir, "bciall_infestation_histories.csv"))
```

## 1.7 Build the focal-tree master table

```{r}
#| label: master-table

# Merge infestation histories with dendrometer metadata, fieldwork records,
# and TLS-derived XY coordinates to produce the all_info_focal_trees table.
# This is the primary input for the ITSMe metric extraction step below.

# One dendrometer row per tree (most recent visit)
dendro_latest <- bciall2 |>
  filter(!is.na(tag6)) |>
  arrange(tag6, desc(date)) |>
  group_by(tag6) |>
  slice_head(n = 1) |>
  ungroup() |>
  select(tag6, calc.dbh, dendroDiam, dendroHt, paintDiam, paintHt,
         crown, illumination, DendroID, measure, date)

# TLS-aligned coordinates produced by the ply-matching step (section 2)
tls_all_with_aligned_coords <- read_csv(
  # insert path to tls_all_with_aligned_coords.csv
  show_col_types = FALSE)

tls_keep <- tls_all_with_aligned_coords |>
  mutate(tag = as.character(tag)) |>
  select(tag, px, py, fid, quadrnt, x_aligned, y_aligned) |>
  distinct(tag, .keep_all = TRUE)

# Join everything onto the infestation histories table.
# The 'full' version (all_info_focal_trees2) retains all overlapping trees
# regardless of species; the filtered version restricts to focal species
# with ≥10 individuals for H2b analyses.
bciall_histories_full <- build_histories(bciall_wide_filled, yrs = 2011:2024)

all_info_focal_trees <- bciall_histories_full |>
  rename(species_code = species) |>
  left_join(dendro_latest, by = "tag6") |>
  left_join(fieldwork_clean_v2 |>
              mutate(tag6 = if_else(!is.na(tag),
                                    sprintf("%06d", as.integer(tag)), tag6)),
            by = "tag6", suffix = c("", "_field")) |>
  mutate(tag = coalesce(as.character(tag), tag6)) |>
  left_join(tls_keep, by = "tag", suffix = c("", "_tls")) |>
  relocate(tag6, tag, species_code, species, .before = everything())

message("all_info_focal_trees: ", nrow(all_info_focal_trees),
        " rows, ", n_distinct(all_info_focal_trees$tag6), " unique trees")

write_csv(all_info_focal_trees, file.path(out_dir, "all_info_focal_trees.csv"))
```

---

# 2 TLS Post-Processing — PLY Matching Algorithm

## 2.1 Background

Following individual tree segmentation with RayCloudTools, each detected
object is output as a `.ply` file within a subplot-level folder. To link
these objects to known ForestGEO stem identities, a nearest-neighbour
matching algorithm is applied: the 2D centroid of each `.ply` point cloud
is compared against the georeferenced stem-map coordinates derived from
manual alignment in CloudCompare (translation, rotation, and scale
computed using a subset of large landmark trees as control points per
subplot). Objects with a vertical extent < 10 m are excluded before
matching to remove understorey shrubs and segmentation artefacts. The
four nearest stem-map candidates are retained for each object, and the
correct match is subsequently confirmed manually against DBH and species
identity. This section shows the matching algorithm for a single subplot;
the same code is run for each of the 24 scanned subplots.

## 2.2 PLY matching

```{r}
#| label: ply-matching

library(dplyr)
library(readr)
library(stringr)
library(RANN)
library(Rvcg)

# ── Paths (update for each subplot) ───────────────────────────────────────────
# Folder containing the .ply files output by RayCloudTools for this subplot
ply_root    <- # insert path to RayCloudTools ply output folder for this subplot

# Output directory for copied matched .ply files
dst_dir     <- # insert path to output directory for matched ply files

# Output directory for the matches summary CSV
out_csv_dir <- # insert path to output directory for per-subplot summary CSVs

# The stem-map table with georeferenced XY coordinates and subplot IDs
master_file <- # insert path to final_extraction_map.csv

# ── Parameters ─────────────────────────────────────────────────────────────────
# Change to the number of the subplot being matched
target_subplot        <- 31L           # say change to number of subplot being matched

# How many nearest stem-map candidates to retain per detected object
k_nearest             <- 4L

# Minimum vertical extent (m): objects shorter than this are excluded
# (removes understorey shrubs and segmentation noise)
min_vertical_extent_m <- 10

dir.create(dst_dir,     recursive = TRUE, showWarnings = FALSE)
dir.create(out_csv_dir, recursive = TRUE, showWarnings = FALSE)

# ── Load stem-map data ─────────────────────────────────────────────────────────
master_df <- read_csv(master_file, show_col_types = FALSE)

trees_sub <- master_df |>
  filter(sbpltnm == target_subplot)

if (nrow(trees_sub) == 0) stop("No trees found for subplot: ", target_subplot)

# ── Filter PLY files by vertical extent ───────────────────────────────────────
# Noise objects (labelled -1 by RayCloudTools) are excluded first.
ply_files <- list.files(ply_root, pattern = "\\.ply$",
                        full.names = TRUE, ignore.case = TRUE)
ply_files <- ply_files[!grepl("_-1\\.ply$", basename(ply_files),
                              ignore.case = TRUE)]

message("Filtering ", length(ply_files), " files by vertical extent (≥ ",
        min_vertical_extent_m, " m)...")

filter_ply <- function(f) {
  ply <- try(Rvcg::vcgPlyRead(f, updateNormals = FALSE), silent = TRUE)
  if (inherits(ply, "try-error") || is.null(ply$vb)) return(FALSE)
  z_range <- range(ply$vb[3, ], na.rm = TRUE)
  (z_range[2] - z_range[1]) >= min_vertical_extent_m
}

keep_idx           <- vapply(ply_files, filter_ply, logical(1))
ply_files_filtered <- ply_files[keep_idx]
message("Retained ", length(ply_files_filtered), " candidate trees.")

# ── Nearest-neighbour matching ─────────────────────────────────────────────────
# For each retained .ply, compute the 2D distance from every tree in the
# stem map to the closest point in the .ply point cloud. This fills a
# distance matrix (n_trees × n_plies); rows are sorted to find the k nearest
# .ply candidates for each stem-map entry.

tree_coords <- as.matrix(trees_sub[, c("x_aligned", "y_aligned")])
n_trees     <- nrow(tree_coords)
n_plies     <- length(ply_files_filtered)

dist_mat <- matrix(Inf, nrow = n_trees, ncol = n_plies)

for (i in seq_along(ply_files_filtered)) {
  ply <- Rvcg::vcgPlyRead(ply_files_filtered[i], updateNormals = FALSE)
  pts <- t(ply$vb[1:2, ])                    # XY only
  nn  <- RANN::nn2(data = pts, query = tree_coords, k = 1)
  dist_mat[, i] <- nn$nn.dists[, 1]
}

# ── Compile results ────────────────────────────────────────────────────────────
file_names  <- basename(ply_files_filtered)
k_use       <- min(k_nearest, n_plies)

# For each stem-map tree, record the k nearest .ply files and their distances
match_indices <- t(apply(dist_mat, 1, function(x) order(x)[1:k_use]))

results_list <- lapply(1:k_use, function(k) {
  tibble(
    !!paste0("nearest", k, "_ply")    := file_names[match_indices[, k]],
    !!paste0("nearest", k, "_dist_m") := vapply(
      1:n_trees, function(r) dist_mat[r, match_indices[r, k]], numeric(1))
  )
})

final_df <- bind_cols(trees_sub, bind_cols(results_list))

# Save and copy matched files
out_matches_csv <- file.path(out_csv_dir,
                              paste0("subplot", target_subplot, "_matches.csv"))
write_csv(final_df, out_matches_csv)

selected_plies <- unique(na.omit(as.character(
  unlist(final_df[, grep("_ply$", names(final_df))]))))
src_paths <- ply_files_filtered[basename(ply_files_filtered) %in% selected_plies]
file.copy(from = src_paths, to = dst_dir, overwrite = TRUE)

message("Done. Matches CSV saved to: ", out_matches_csv)
```

---

# 3 Crown Architecture Metric Extraction

## 3.1 Background

Crown metrics — height, crown depth, projected area, and crown volume —
were extracted from each segmented individual tree point cloud using the
ITSMe R package (Terryn et al., 2023). The crown was defined as all
points above the first major branch, with buttress detection applied where
relevant to prevent buttress points being incorrectly assigned to the
crown. Projected crown area is estimated as the area enclosed by a 2D
concave hull (alpha = 2); crown volume as the 3D alpha-shape volume
(alpha = 1). Tree height is the vertical range of the full point cloud;
crown depth is the vertical range of the crown points alone.

The pipeline below processes new files incrementally, skipping trees
that already have entries in the results CSV. This allows the pipeline
to be resumed after interruption.

## 3.2 ITSMe metric extraction

```{r}
#| label: itsme-extraction

library(ITSMe)
library(alphashape3d)
library(dplyr)
library(readr)

options(rgl.useNULL = TRUE)   # suppress rgl window in non-interactive mode

# ── Paths ──────────────────────────────────────────────────────────────────────
pc_folder   <- # insert path to folder containing all individual segmented
               # tree point clouds (.txt files, one per tree)
results_csv <- # insert path to output CSV for ITSMe metrics
               # (ITSMe_tree_metrics_summary_final.csv)

# Helper: return NA instead of throwing for missing outputs
safe_val <- function(x) if (length(x) == 0 || is.null(x)) NA else x

# ── Identify which trees still need processing ─────────────────────────────────
pc_files <- list.files(pc_folder, pattern = "\\.txt$", full.names = TRUE)
pc_names <- basename(pc_files)

if (file.exists(results_csv)) {
  old_results  <- read_csv(results_csv, show_col_types = FALSE)
  done         <- if ("filename" %in% names(old_results))
    unique(old_results$filename) else character(0)
  files_to_run <- pc_files[!(pc_names %in% done)]
} else {
  old_results  <- NULL
  files_to_run <- pc_files
}

cat("Total trees in folder:", length(pc_files), "\n")
cat("New trees to process:", length(files_to_run), "\n")
if (length(files_to_run) == 0) { cat("Nothing to do.\n"); stop() }

# ── Process each new tree ──────────────────────────────────────────────────────
results_list <- vector("list", length(files_to_run))

for (i in seq_along(files_to_run)) {
  f <- files_to_run[i]
  cat("Processing:", basename(f), "(", i, "/", length(files_to_run), ")\n")

  pc_tree <- read_tree_pc(path = f)

  # DBH and diameter above buttress (DAB); DAB height is used to set the
  # crown base height where buttresses would otherwise inflate crown depth
  out_dbh   <- dbh_pc(pc = pc_tree, plot = FALSE)
  out_dab   <- dab_pc(pc = pc_tree, plot = FALSE)
  buttress_h <- out_dab$h

  # Full tree height
  H_out <- tree_height_pc(pc = pc_tree)

  # Crown classification: if a buttress taller than 1.5 m is detected,
  # set the crown base accordingly; otherwise use the default algorithm
  C_out <- if (!is.null(buttress_h) && !is.na(buttress_h) && buttress_h > 1.5)
    classify_crown_pc(pc = pc_tree, minheight = buttress_h,
                      buttress = TRUE, plot = FALSE)
  else
    classify_crown_pc(pc = pc_tree, plot = FALSE)

  crown_pc   <- C_out$crownpoints
  crownDepth <- tree_height_pc(pc = crown_pc)             # crown depth
  pca_out    <- projected_area_pc(pc = crown_pc, plot = FALSE) # 2D alpha hull

  # Alpha-shape crown volume — handle version differences in return structure
  cvol_out <- tryCatch({
    raw <- alpha_volume_pc(pc = crown_pc, plot = FALSE)
    if (is.list(raw)) raw$av else raw
  }, error = function(e) NA)

  results_list[[i]] <- data.frame(
    filename     = basename(f),
    dbh          = safe_val(out_dbh$dbh),
    dab          = safe_val(out_dab$dab),
    buttress_h   = safe_val(out_dab$h),
    height       = safe_val(H_out$h),
    crown_depth  = safe_val(crownDepth$h),
    pa           = safe_val(pca_out),         # projected (crown) area in m²
    crown_vol    = safe_val(cvol_out),         # alpha-shape volume in m³
    stringsAsFactors = FALSE
  )
}

# ── Append to existing results, deduplicate, and save ─────────────────────────
new_results <- do.call(rbind, results_list)

if (!is.null(old_results)) {
  all_cols <- union(names(old_results), names(new_results))
  for (nm in setdiff(all_cols, names(old_results)))  old_results[[nm]]  <- NA
  for (nm in setdiff(all_cols, names(new_results)))  new_results[[nm]]  <- NA
  updated <- bind_rows(old_results[, all_cols], new_results[, all_cols])
} else {
  updated <- new_results
}

updated <- updated |>
  arrange(filename) |>
  distinct(filename, .keep_all = TRUE)

write_csv(updated, results_csv)
cat("Done. Rows in results table:", nrow(updated), "\n")
```

---

# 4 Plant Area Index Estimation

## 4.1 Background

Plant area index (PAI) quantifies the one-sided projected area of all
plant material per unit ground area. PAI was estimated from TLS point
clouds using the voxel-based vertical-slicing approach of Flynn et al.
(2023), with a voxel resolution of 0.10 m and a leaf-inclination
correction factor of 1.1. Following leaf-wood separation applied to each
individual tree point cloud, two PAI values were computed per tree:

- **Host PAI** (tree-only): from the crown point cloud, clipped to all
  points above the first branch. This captures host foliage density
  independently of any associated liana material.
- **Combined PAI** (host + liana): an alpha-shape hull (α = 100)
  constructed around the host leaf points was used to query the
  co-registered subplot-level point cloud, capturing all vegetation
  within the crown volume — host foliage, host wood, and liana material.

A sensitivity analysis was conducted on a stratified subsample of 60 trees
(20 per liana load category: none, light, heavy) to confirm stability of
PAI estimates to the choice of alpha value and voxel resolution.

## 4.2 Separate leaf and wood points

```{r}
#| label: leaf-wood-separation

library(data.table)
library(fs)

# Each input file contains four columns: X, Y, Z, class
# (class 0 = leaf, class 1 = wood), output by the leaf-wood separator.
input_tree_dir  <- # insert path to folder containing leaf-wood separated tree files
leaf_output_dir <- # insert path to output folder for leaf points
wood_output_dir <- # insert path to output folder for wood points

dir_create(leaf_output_dir)
dir_create(wood_output_dir)

tree_files <- dir_ls(input_tree_dir, glob = "*.txt")
message("Separating ", length(tree_files), " trees...")

for (f in tree_files) {
  tree_id  <- path_ext_remove(basename(f))
  dat      <- fread(f, col.names = c("X", "Y", "Z", "class"))
  leaf_pts <- dat[class == 0, .(X, Y, Z)]
  wood_pts <- dat[class == 1, .(X, Y, Z)]
  if (nrow(leaf_pts) > 0)
    fwrite(leaf_pts, file.path(leaf_output_dir, paste0(tree_id, "_leaf.txt")),
           sep = "\t", col.names = FALSE)
  if (nrow(wood_pts) > 0)
    fwrite(wood_pts, file.path(wood_output_dir, paste0(tree_id, "_wood.txt")),
           sep = "\t", col.names = FALSE)
  rm(dat, leaf_pts, wood_pts); gc()
}
message("Leaf-wood separation complete.")
```

## 4.3 Organise files by subplot

```{r}
#| label: organise-by-subplot

library(data.table); library(fs); library(dplyr); library(readr)

# Leaf and tree files are grouped into per-subplot subdirectories so that
# each subplot's plot-level point cloud only needs to be loaded once during
# the combined extraction step below.
source_tree_dir    <- # insert path to folder containing all individual tree files
target_tree_by_sub <- # insert path to output folder for subplot-organised tree files
source_leaf_dir    <- # insert path to leaf points folder (from step 4.2)
target_leaf_by_sub <- # insert path to output folder for subplot-organised leaf files
metadata_csv       <- # insert path to metadata CSV with columns: Tag, Subplot number

dir_create(target_tree_by_sub)
dir_create(target_leaf_by_sub)

new_trees <- read_csv(metadata_csv)

# Move tree files into subplot subdirectories
for (tf in dir_ls(source_tree_dir, glob = "*.txt")) {
  tag_id   <- as.numeric(gsub(".*_([0-9]+)\\.txt$", "\\1", basename(tf)))
  sub_info <- new_trees |> filter(Tag == tag_id)
  if (nrow(sub_info) == 0) next
  sub_folder <- file.path(target_tree_by_sub,
                           paste0("subplot_", sub_info$`Subplot number`[1]))
  dir_create(sub_folder)
  file_move(tf, file.path(sub_folder, basename(tf)))
}

# Copy leaf files into subplot subdirectories (copy not move, to preserve originals)
for (lf in dir_ls(source_leaf_dir, glob = "*.txt")) {
  tag_id   <- as.numeric(gsub(".*_([0-9]+)_leaf\\.txt$", "\\1", basename(lf)))
  sub_info <- new_trees |> filter(Tag == tag_id)
  if (nrow(sub_info) == 0) next
  sub_folder <- file.path(target_leaf_by_sub,
                           paste0("subplot_", sub_info$`Subplot number`[1]))
  dir_create(sub_folder)
  file_copy(lf, file.path(sub_folder, basename(lf)), overwrite = TRUE)
}
message("Files organised by subplot.")
```

## 4.4 Extract combined host + liana point clouds

```{r}
#| label: combined-extraction

library(data.table); library(fs); library(alphashape3d)

# For each tree, an alpha-shape hull (α = 100) is fitted to the host leaf
# points and used to query the full subplot-level point cloud. This captures
# all plant material within the crown volume — host foliage, host wood, and
# any liana points — providing the combined PAI input. α = 100 approximates
# a convex hull, ensuring the entire crown volume is captured; sensitivity
# to this choice is evaluated in section 5.
plot_level_dir   <- # insert path to folder containing per-subplot downsampled
                    # point clouds (files named subplot{N}_downsample.txt)
leaf_by_subplot  <- # insert path to subplot-organised leaf files folder
combined_out_dir <- # insert path to output folder for combined host+liana files

dir_create(combined_out_dir)

subplot_folders <- dir_ls(leaf_by_subplot, type = "directory")

for (fld in subplot_folders) {
  sub_num  <- gsub("[^0-9]", "", basename(fld))
  sub_file <- file.path(plot_level_dir,
                         paste0("subplot", sub_num, "_downsample.txt"))
  if (!file_exists(sub_file)) {
    message("Skipping subplot ", sub_num, " — plot cloud not found."); next
  }
  message("Loading subplot ", sub_num, "...")
  subplot_pc <- fread(sub_file, select = 1:3,
                      col.names = c("X", "Y", "Z"))

  for (lf in dir_ls(fld, glob = "*.txt")) {
    tree_id  <- sub("_leaf\\.txt$", "", basename(lf))
    leaf_pts <- fread(lf, select = 1:3, col.names = c("X", "Y", "Z"))
    leaf_mat <- as.matrix(unique(leaf_pts))
    if (nrow(leaf_mat) < 5) next

    ashape <- try(ashape3d(leaf_mat, alpha = 100), silent = TRUE)
    if (inherits(ashape, "try-error")) next

    # Crop the subplot cloud to a 0.2 m buffered bounding box before the
    # (slower) point-in-hull check — substantially reduces computation time
    tree_box <- subplot_pc[
      X %between% (range(leaf_pts$X) + c(-0.2, 0.2)) &
      Y %between% (range(leaf_pts$Y) + c(-0.2, 0.2)) &
      Z %between% (range(leaf_pts$Z) + c(-0.2, 0.2))]
    if (nrow(tree_box) == 0) next

    inside_idx   <- inashape3d(ashape, indexAlpha = 1,
                                p = as.matrix(tree_box))
    combined_pts <- tree_box[inside_idx >= 1]

    if (nrow(combined_pts) > 0)
      fwrite(combined_pts,
             file.path(combined_out_dir, paste0(tree_id, "_combined.txt")),
             sep = "\t")
  }
  rm(subplot_pc); gc()
}
message("Combined files saved to: ", combined_out_dir)
```

## 4.5 Clip tree point clouds to crown only

```{r}
#| label: crown-clipping

library(data.table); library(fs); library(dplyr); library(readr)

# For host PAI estimation, each full tree point cloud is clipped to retain
# only points at or above the first major branch. The crown base height
# is derived from ITSMe metrics as total height − crown depth.
tree_dir      <- # insert path to folder containing full individual tree files
crown_dir     <- # insert path to output folder for crown-only point clouds
itsme_metrics <- # insert path to ITSMe_tree_metrics_summary_final.csv

dir_create(crown_dir)

itsme <- read_csv(itsme_metrics, show_col_types = FALSE) |>
  mutate(tag = as.numeric(gsub(".*_([0-9]+)\\.txt$", "\\1", filename)))

for (tf in dir_ls(tree_dir, glob = "*.txt")) {
  file_name <- basename(tf)
  tag_id    <- as.numeric(gsub(".*_([0-9]+)\\.txt$", "\\1", file_name))
  met       <- filter(itsme, tag == tag_id)
  if (nrow(met) == 0) { message("!! Skipping ", file_name, " — not in ITSMe."); next }

  dat <- fread(tf, col.names = c("X", "Y", "Z", "class"))
  if (nrow(dat) == 0) next

  z_cutoff  <- min(dat$Z) + (met$height[1] - met$crown_depth[1])
  crown_pts <- dat[Z >= z_cutoff]

  if (nrow(crown_pts) == 0) {
    message("!! No crown points for ", file_name); next
  }
  fwrite(crown_pts, file.path(crown_dir, file_name), sep = "\t",
         col.names = FALSE)
  rm(dat, crown_pts); gc()
}
message("Crown files saved to: ", crown_dir)
```

## 4.6 PAI calculation

```{r}
#| label: pai-calculation

library(data.table); library(VoxR); library(raster)

# Core PAI function: voxelises the point cloud at 0.10 m resolution and
# sums the per-slice contact frequency, scaled by a leaf-inclination
# correction factor of 1.1 (Flynn et al., 2023).
# PAI = Σ_k [ 1.1 × (occupied XY cells in slice k) / (total crown footprint cells) ]

compute_PAI <- function(file, tree_id, res = 0.1) {
  pts <- fread(file, select = 1:3, col.names = c("x", "y", "z"))
  if (nrow(pts) == 0)
    return(data.frame(tree = tree_id, PAI = NA,
                      slices_occupied = 0L, crown_height_m = NA))
  vox <- VoxR::vox(pts, res = res, full.grid = FALSE)
  if (nrow(vox) == 0)
    return(data.frame(tree = tree_id, PAI = NA,
                      slices_occupied = 0L, crown_height_m = NA))
  rt           <- rasterFromXYZ(vox[, 1:3, drop = FALSE])
  ncells_total <- ncell(rt)
  z_seq        <- seq(min(vox$z), max(vox$z), by = res)
  pai_vec      <- numeric(length(z_seq))
  for (i in seq_along(z_seq)) {
    slice <- vox[abs(vox$z - z_seq[i]) <= (res / 2), , drop = FALSE]
    if (nrow(slice) > 0)
      pai_vec[i] <- 1.1 *
        length(unique(cellFromXY(rt, slice[, 1:2, drop = FALSE]))) /
        ncells_total
  }
  data.frame(tree = tree_id, PAI = sum(pai_vec),
             slices_occupied = sum(pai_vec > 0),
             crown_height_m  = length(z_seq) * res)
}

# ── Host-tree PAI ───────────────────────────────────────────────────────────────
crown_dir  <- # insert path to crown folder (from step 4.5)
host_out   <- # insert path for output CSV: PAI_crown_results.csv

crown_files <- list.files(crown_dir, pattern = "\\.txt$", full.names = TRUE)
tree_ids    <- sub("\\.txt$", "", basename(crown_files))

write.table(data.frame(tree=character(), PAI=numeric(),
                        slices_occupied=integer(), crown_height_m=numeric()),
            host_out, sep=",", row.names=FALSE, col.names=TRUE)
for (k in seq_along(tree_ids)) {
  message("Host PAI: ", tree_ids[k], " (", k, "/", length(tree_ids), ")")
  write.table(compute_PAI(crown_files[k], tree_ids[k]),
              host_out, sep=",", row.names=FALSE, col.names=FALSE, append=TRUE)
}

# ── Combined PAI ────────────────────────────────────────────────────────────────
combined_dir <- # insert path to combined host+liana files folder (from step 4.4)
comb_out     <- # insert path for output CSV: PAI_combined_results.csv

combined_files <- list.files(combined_dir, pattern="_combined\\.txt$",
                             full.names=TRUE)
comb_ids       <- gsub(".*_([0-9]+)_combined\\.txt$", "\\1",
                       basename(combined_files))

write.table(data.frame(tree=character(), PAI=numeric(),
                        slices_occupied=integer(), crown_height_m=numeric()),
            comb_out, sep=",", row.names=FALSE, col.names=TRUE)
for (k in seq_along(comb_ids)) {
  message("Combined PAI: ", comb_ids[k], " (", k, "/", length(comb_ids), ")")
  write.table(compute_PAI(combined_files[k], comb_ids[k]),
              comb_out, sep=",", row.names=FALSE, col.names=FALSE, append=TRUE)
}
```

## 4.7 Compile final PAI results table

```{r}
#| label: compile-pai

library(dplyr); library(readr); library(fs)

host_out    <- # insert path to PAI_crown_results.csv (from step 4.6)
comb_out    <- # insert path to PAI_combined_results.csv (from step 4.6)
pai_out_dir <- # insert path to output folder for final results
out_file    <- file.path(pai_out_dir, "PAI_h1.2_results.csv")

dir_create(pai_out_dir)

df_crown <- read_csv(host_out, show_col_types = FALSE) |>
  rename(file_name = tree, PAI_tree_only = PAI) |>
  mutate(tag = as.numeric(sub(".*_([0-9]+).*", "\\1", file_name)))

df_comb <- read_csv(comb_out, show_col_types = FALSE) |>
  rename(PAI_combined = PAI) |>
  mutate(tag = as.numeric(sub("^[A-Za-z0-9]{6}_([0-9]{3,6})$", "\\1", tree))) |>
  select(tag, PAI_combined)

# Report any trees present in host-only but not combined (excluded from merge)
missing <- setdiff(df_crown$tag, df_comb$tag)
if (length(missing) > 0) {
  message(length(missing), " tree(s) in host-only but not combined — excluded:")
  print(filter(df_crown, tag %in% missing) |> select(file_name, tag, PAI_tree_only))
}

df_final <- inner_join(df_crown, df_comb, by = "tag")
message("Final PAI table: ", nrow(df_final), " trees")
write_csv(df_final, out_file)
message("Saved: ", out_file)
```

---

# 5 PAI Sensitivity Analyses

## 5.1 Background

PAI is sensitive to both the alpha value used to define the liana
extraction hull and the voxel resolution. A stratified subsample of 60
trees (20 per liana load category: none, light, heavy) was used to
assess stability across a range of both parameters. A voxel size of
0.10 m and an alpha value of 100 were selected: PAI stabilised above
α = 50 in the alpha sensitivity analysis, and 0.10 m sat within the
stable region of the voxel sensitivity curve — small enough to resolve
individual leaves but large enough to avoid noise artefacts from empty
voxels (Wei et al., 2020).

## 5.2 Select trees for sensitivity analysis

```{r}
#| label: sensitivity-selection

library(dplyr); library(readr)

# Stratified random sample: 20 trees from each liana load category.
# df_final (from step 4.7) and fieldwork metadata must be in the environment.

full_dataset <- df_final |>
  inner_join(fieldwork, by = "tag")

sensitivity_selection <- full_dataset |>
  mutate(liana_cat = case_when(
    lianas_2019 == 0                    ~ "none",
    lianas_2019 > 0 & lianas_2019 <= 50 ~ "light",
    lianas_2019 > 50                    ~ "heavy",
    TRUE ~ NA_character_)) |>
  filter(!is.na(PAI_combined), !is.na(liana_cat)) |>
  group_by(liana_cat) |>
  slice_sample(n = 20) |>
  ungroup()

print(count(sensitivity_selection, liana_cat))

sensitivity_tags_file <- # insert path for output CSV: sensitivity_analysis_tags.csv
write_csv(sensitivity_selection, sensitivity_tags_file)
```

## 5.3 Alpha-shape parameter sensitivity

```{r}
#| label: alpha-sensitivity

library(data.table); library(fs); library(alphashape3d)
library(VoxR); library(raster); library(dplyr)

plot_level_dir <- # insert path to subplot-level downsampled point clouds
leaf_dir       <- # insert path to leaf-cleaned files folder
wood_dir       <- # insert path to wood points folder
sens_out_dir   <- # insert path to output folder for alpha sensitivity files
sens_results   <- # insert path for output CSV: sensitivity_results_final.csv

ALPHA_VALS <- c(1, 25, 50, 100, 125)
BUFFER     <- 0.2
target_trees <- as.data.table(sensitivity_selection)

# For each alpha value, re-extract combined point clouds and recalculate PAI
results <- list()

for (current_alpha in ALPHA_VALS) {
  version_dir <- file.path(sens_out_dir, paste0("alpha_", current_alpha))
  dir_create(version_dir)
  message(">>> Alpha: ", current_alpha)

  for (i in seq_len(nrow(target_trees))) {
    file_id   <- target_trees$file_name[i]
    sub_id    <- target_trees$sbpltnm[i]
    leaf_path <- file.path(leaf_dir, paste0(file_id, "_leaf.txt"))
    wood_path <- file.path(wood_dir, paste0(file_id, "_wood.txt"))
    sub_file  <- file.path(plot_level_dir,
                            paste0("subplot", sub_id, "_downsample.txt"))
    out_path  <- file.path(version_dir,
                            paste0(file_id, "_alpha", current_alpha, ".txt"))
    if (file_exists(out_path) || !file_exists(leaf_path)) next

    leaf_pts   <- fread(leaf_path, select = 1:3, col.names = c("X","Y","Z"))
    ashape     <- try(ashape3d(as.matrix(leaf_pts), alpha = current_alpha),
                      silent = TRUE)
    if (inherits(ashape, "try-error")) next

    subplot_pc <- fread(sub_file, select = 1:3, col.names = c("X","Y","Z"))
    tree_box   <- subplot_pc[
      X %between% (range(leaf_pts$X) + c(-BUFFER, BUFFER)) &
      Y %between% (range(leaf_pts$Y) + c(-BUFFER, BUFFER)) &
      Z %between% (range(leaf_pts$Z) + c(-BUFFER, BUFFER))]

    inside_idx   <- inashape3d(ashape, indexAlpha = 1,
                                p = as.matrix(tree_box[, 1:3]))
    foliage_only <- tree_box[inside_idx >= 1]

    # Remove wood points to ensure PAI reflects leaf area only
    if (file_exists(wood_path)) {
      wood_pts     <- fread(wood_path, select = 1:3, col.names = c("X","Y","Z"))
      setkey(foliage_only, X, Y, Z); setkey(wood_pts, X, Y, Z)
      foliage_only <- foliage_only[!wood_pts]
    }
    if (nrow(foliage_only) > 0)
      fwrite(foliage_only, out_path, sep = "\t")
    rm(subplot_pc, leaf_pts, ashape, tree_box, foliage_only); gc()
  }

  # Calculate PAI for this alpha version
  for (f in dir_ls(version_dir, glob = "*.txt")) {
    file_id <- sub(paste0("_alpha", current_alpha, "\\.txt$"), "", basename(f))
    pts     <- fread(f, select = 1:3, col.names = c("x","y","z"))
    if (nrow(pts) == 0) next
    vox          <- VoxR::vox(pts, res = 0.1, full.grid = FALSE)
    rt           <- rasterFromXYZ(vox[, 1:3, drop = FALSE])
    ncells_total <- ncell(rt)
    z_seq        <- seq(min(vox$z), max(vox$z), by = 0.1)
    pai_val <- sum(sapply(z_seq, function(z0) {
      slice <- vox[abs(vox$z - z0) <= 0.05]
      if (nrow(slice) == 0) return(0)
      1.1 * length(unique(cellFromXY(rt, slice[, 1:2, drop = FALSE]))) /
        ncells_total
    }))
    results[[paste0(file_id, "_a", current_alpha)]] <-
      data.table(tree_id = file_id, alpha = current_alpha, pai = pai_val)
  }
}

sens_dt <- bind_rows(results)
fwrite(sens_dt, sens_results)
message("Alpha sensitivity results saved.")
```

### Alpha sensitivity figures

```{r}
#| label: alpha-sensitivity-figures
#| eval: true   # these can be run once sens_dt is loaded

library(ggplot2); library(dplyr)

figs_dir <- # insert path to output folder for figures

RESCALE_FACTOR <- 12     # correction factor used in main analysis

sens_dt_scaled <- sens_dt |>
  mutate(pai_corrected = pai * RESCALE_FACTOR)

# Remove trees with implausibly large PAI (likely segmentation artefacts)
outliers <- sens_dt_scaled |>
  group_by(tree_id) |>
  summarise(max_pai = max(pai_corrected, na.rm = TRUE)) |>
  filter(max_pai > 15) |>
  pull(tree_id)

sens_clean <- filter(sens_dt_scaled, !tree_id %in% outliers)

sens_theme <- theme_classic(base_size = 12) +
  theme(axis.title = element_text(size = 12),
        axis.text  = element_text(size = 11))

# Individual tree trajectories — shows tree-level stability
p_alpha_trees <- ggplot(sens_clean,
                         aes(x = alpha, y = pai_corrected, group = tree_id)) +
  geom_line(colour = "grey50", alpha = 0.5, linewidth = 0.4) +
  geom_point(colour = "grey30", size = 1.2) +
  geom_vline(xintercept = 100, linetype = "dashed",
             colour = "firebrick", linewidth = 0.7) +
  annotate("text", x = 103,
           y = max(sens_clean$pai_corrected, na.rm = TRUE) * 0.97,
           label = "\u03b1 = 100\n(used in analysis)",
           hjust = 0, colour = "firebrick", size = 3.2) +
  scale_x_continuous(breaks = c(1, 25, 50, 100, 125)) +
  scale_y_continuous(limits = c(0, NA)) +
  sens_theme +
  labs(x = "Alpha value",
       y = expression("PAI (m"^2~"m"^-2~")"),
       title = "Sensitivity of PAI to alpha-shape parameter")

# Mean ± SD ribbon — shows population-level stability
sens_summary <- sens_clean |>
  group_by(alpha) |>
  summarise(mean_pai = mean(pai_corrected, na.rm = TRUE),
            sd_pai   = sd(pai_corrected,   na.rm = TRUE), .groups = "drop")

p_alpha_mean <- ggplot(sens_summary, aes(x = alpha, y = mean_pai)) +
  geom_ribbon(aes(ymin = mean_pai - sd_pai, ymax = mean_pai + sd_pai),
              fill = "steelblue", alpha = 0.25) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 2.5) +
  geom_vline(xintercept = 100, linetype = "dashed",
             colour = "firebrick", linewidth = 0.7) +
  scale_x_continuous(breaks = c(1, 25, 50, 100, 125)) +
  scale_y_continuous(limits = c(0, NA)) +
  sens_theme +
  labs(x = "Alpha value",
       y = expression("Mean PAI \u00b1 1 SD (m"^2~"m"^-2~")"),
       title = "Mean PAI across 60 trees by alpha value")

ggsave(file.path(figs_dir, "alpha_sensitivity_all_trees.png"),
       p_alpha_trees, width = 8, height = 5, dpi = 300)
ggsave(file.path(figs_dir, "alpha_sensitivity_summary.png"),
       p_alpha_mean,  width = 7, height = 5, dpi = 300)
```

## 5.4 Voxel resolution sensitivity

```{r}
#| label: voxel-sensitivity

library(data.table); library(dplyr); library(VoxR); library(raster)

crown_dir      <- # insert path to crown folder (from step 4.5)
voxel_out_file <- # insert path for output CSV: VOXEL_sensitivity_results.csv

VOXEL_SIZES    <- c(0.025, 0.05, 0.1, 0.15, 0.2)
focal_tree_ids <- sensitivity_selection$file_name

write.table(
  data.frame(tree = character(), voxel_res = numeric(), PAI = numeric(),
             slices_occupied = integer(), crown_height_m = numeric()),
  voxel_out_file, sep = ",", row.names = FALSE, col.names = TRUE)

for (v_res in VOXEL_SIZES) {
  message("Voxel size: ", v_res, " m")
  for (t_id in focal_tree_ids) {
    crown_path <- file.path(crown_dir, paste0(t_id, ".txt"))
    if (!file.exists(crown_path)) next
    pts <- fread(crown_path, select = 1:3, col.names = c("x","y","z"))
    if (nrow(pts) == 0) next
    vox          <- VoxR::vox(pts, res = v_res, full.grid = FALSE)
    rt           <- rasterFromXYZ(vox[, 1:3, drop = FALSE])
    ncells_total <- ncell(rt)
    z_seq        <- seq(min(vox$z), max(vox$z), by = v_res)
    pai_vec      <- numeric(length(z_seq))
    for (i in seq_along(z_seq)) {
      slice <- vox[abs(vox$z - z_seq[i]) <= (v_res / 2), , drop = FALSE]
      if (nrow(slice) > 0)
        pai_vec[i] <- 1.1 *
          length(unique(cellFromXY(rt, slice[, 1:2, drop = FALSE]))) /
          ncells_total
    }
    write.table(
      data.frame(tree = t_id, voxel_res = v_res, PAI = sum(pai_vec),
                 slices_occupied = sum(pai_vec > 0),
                 crown_height_m  = length(z_seq) * v_res),
      voxel_out_file, sep = ",", row.names = FALSE,
      col.names = FALSE, append = TRUE)
  }
}
message("Voxel sensitivity analysis complete.")
```

### Voxel sensitivity figure

```{r}
#| label: voxel-sensitivity-figure

library(ggplot2); library(readr)

voxel_results <- read_csv(voxel_out_file, show_col_types = FALSE)

p_voxel <- ggplot(voxel_results,
                   aes(x = voxel_res, y = PAI, group = tree)) +
  geom_line(colour = "grey50", alpha = 0.5, linewidth = 0.4) +
  geom_point(colour = "grey30", size = 1.2) +
  geom_vline(xintercept = 0.10, linetype = "dashed",
             colour = "firebrick", linewidth = 0.7) +
  annotate("text", x = 0.105,
           y = max(voxel_results$PAI, na.rm = TRUE) * 0.97,
           label = "0.10 m\n(used in analysis)",
           hjust = 0, colour = "firebrick", size = 3.2) +
  scale_x_continuous(breaks = c(0.025, 0.05, 0.10, 0.15, 0.20)) +
  scale_y_continuous(limits = c(0, NA)) +
  sens_theme +
  labs(x = "Voxel size (m)",
       y = expression("PAI (m"^2~"m"^-2~")"),
       title = "Sensitivity of PAI to voxel resolution")

ggsave(file.path(figs_dir, "voxel_sensitivity.png"),
       p_voxel, width = 8, height = 5, dpi = 300)
```

---

```{r}
#| label: session-info
#| code-fold: false
sessionInfo()
```
